---
title: "glinski_stemflow"
output: html_document
---
tested hypotheses
============
Test to see whether field observations of stemflow concentrations are greater that forest throughfall concentrations (or vice-versa). Samples are paired across 3 locations and 18 different sampling dates. Therefore 54 observations for each of 32 chemicals. However, many samples are non-detected.

Non-parametric Wilcoxon Rank Sum test implemented. Permutation tests also possible. Test requiring distribution assumptions (e.g., t-test) not practical due to non-detects and data skewness.

This page is published at: http://rpubs.com/puruckertom/glinski_stemflow2

figures needed
============

done:

- scatterplot of stemflow concentration versus throughfall concentrations (fig #6)
- figure 6 as jpg (add units to x and y - ug/L)

still needed:

- a single plot stacked surface water concentration versus time for all sites (all 10 tifton pond sites- metolachlor, tebuconazole) figure 3 redone
- figure 4 stacked clustered surface water (4L samples) by site*month (Ponds_4L.csv)
- figure 5 is stemflow and throughfall for tebuconazole and metolachlor (3 sites each), time series (like figure 3)
- a single plot concentration versus time plots for stemflow and throughfall by site, 6 different symbol/color/filled) figure 5 redone (metolachlor, tebuconazole))
- signs versus time plot to support wrs tests
- signs versus rainfall intensity (cumulative rainfall) to support wrs tests

possible additional things to do
============
- wrs tests for stemflow greater than throughfall by date
- wrs tests for stemflow greater than throughfall by site (metolachlor) 
- signs versus volume(s)
- implement permutation tests
- test all 32 chemicals at once rather than one at a time (done)
- effects estimation approaches


general configuration
=====================
Install and load supporting libraries.
```{r eval=TRUE, echo=FALSE}
print(Sys.info()[4])

R.Version()$version.string
library(rmarkdown, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(knitr, quietly = TRUE, warn.conflicts = FALSE)
library(ggplot2)
print("list of loaded packages: ")
print((.packages()))
```

import experimental exposure and dehydration data
=================================
Load csv file with experimental dehydration data. The below may return false but still be OK if rstudio does not have privileges to data directory (e.g., attached drive).

```{r eval=TRUE, echo=FALSE}
if(Sys.info()[4]=="DONNA-PC"){
  stemflow.root <- path.expand("C:/Users/Donna/Documents/Github/glinski_stemflow/")
}

if(Sys.info()[4]=="stp-air-3.local" || Sys.info()[4]=="stp-air.local" || Sys.info()[4]=="stp-air"){
  stemflow.root <- path.expand("~/git/glinski_stemflow/")
}

if(Sys.info()[4]=="DZ2626UTPURUCKE"){
  stemflow.root <- "d:/git/glinski_stemflow/"
}
print(paste("Root directory location: ", stemflow.root, sep=""))

stemflow.csv.in <- paste(stemflow.root, "csv_in/", sep="")
stemflow.csv.out <- paste(stemflow.root, "csv_out/", sep="")
stemflow.graphics <- paste(stemflow.root, "graphics/", sep="")
stemflow.tables <- paste(stemflow.root, "tables/", sep="")

#maybe ok even if remote drive returns false
print(paste("check to see if R can access files OK: ", file.exists(stemflow.csv.in), sep = ""))

#import tifton data
tifton <- read.table(paste(stemflow.csv.in,"tifton2015.csv",sep=""), header = TRUE, sep = ",")
#View(tifton)

#import tifton data
rainwater <- read.table(paste(stemflow.csv.in,"rainwater_amount.csv",sep=""), header = TRUE, sep = ",")


```

tifton data structure
====================================
Check out structure of imported data sets. sf v tf (32 chemicals, 3 paired sites, 52 storm events), pondwater (10 samples * 12 months)

```{r eval=TRUE, echo=TRUE}
summary(tifton)
colnames(tifton)
chems <- unique(tifton$Compound)
chems
media <- unique(tifton$Type)
media
#coerce concentrations to numeric because NAs
tifton$Conc <- as.numeric(tifton$Conc)
min(tifton$Conc,na.rm=TRUE)
max(tifton$Conc,na.rm=TRUE)

colnames(rainwater)
colnames(rainwater)[1] <- 'Date'

summary(rainwater)
unique(rainwater$Date)
#chem_merge <- merge(chem_merge, rainwater, by="Date")
```

wilcox stat testing
====================================
Run Wilcoxon rank sum test on throughfall versus stemflow for each individual chemical that has detected values. There are three different commands in R that calculate
the Wilcoxon signed-rank test; wilcox.test, wilcox.exact, and
wilcoxsign_test. They will each yield a different  p-value.

The nonparametric Wilcoxon signed-rank test seems to be the most
appropriate test for these data. There are two different methods to
calculate the signed-rank test. The first is by Wilcoxon (1945), who
discards any tied data and then calculates the signed ranks. The second
method incorporates tied values in the ranking procedure (see J.W. Pratt,
1959, Remarks on zeros and ties in the Wilcoxon signed rank procedure:
Journal of the American Statistical Association, Vol. 54, No. 287, pp.
655-667). There are two commands in R that calculate the original method
by Wilcoxon, wilcox.test and wilcoxsign_test (make sure
to include the argument "zero.method = c("Wilcoxon")"). There are two
other commands in R that incorporate ties in the signed-rank test,
wilcox.exact and wilcoxsign_test (make sure to include the
argument"zero.method = c("Pratt")").

1. wilcox.test() in the stats package
This function automatically switches to using a Normal
approximation when there are ties in the data:

wilcox.test(x, y, paired=TRUE)$p.value

(You can suppress the warning (due to ties) by specifying
the argument 'exact=FALSE'.)

This function also uses a continuity correction unless
told not to:

wilcox.test(x, y, paired=TRUE, correct=FALSE)$p.value

2. wilcox.exact() in pkg exactRankTests
This function can handle ties (using the "Wilcoxon" method)
with an 'exact' calculation:

wilcox.exact(x, y, paired=TRUE)$p.value

If we want the Normal approximation:

wilcox.exact(x, y, paired=TRUE, exact=FALSE)$p.value

3. wilcoxsign_test() in pkg coin
This is the most comprehensive of these functions.
It is also the only one that offers the "Pratt" method
of handling ties. It will default to this method and
a Normal approximation:

pvalue(wilcoxsign_test(x ~ y))

pvalue(wilcoxsign_test(x ~ y, zero.method="Pratt",
         distribution="asympt"))

You can get the results from wilcox.exact() with

  pvalue(wilcoxsign_test(x ~ y, zero.method="Wilcoxon",
         distribution="asympt"))

and

  pvalue(wilcoxsign_test(x ~ y, zero.method="Wilcoxon",
         dist="exact"))

For the moment we are using wilcox.test with ties (paired = TRUE, alternative = 'greater', exact = FALSE) but the test is not being
run unless there are at least 4 detected values for a chemical.


chemical level analyses
====================================
```{r eval=TRUE, echo=TRUE}
chems_we_care_about <- c('Metolachlor', 'Tebuconazole', '2-Phenylphenol', 'Atrazine', 'Biphenyl', 'DEA', 'Flutolanil')
for(chem in chems_we_care_about){
  print('=======================================================')
  print(chem)
  chem_data <- tifton[which(tifton$Compound==chem),]
  chem_sf <- chem_data[which(chem_data$Type=='SF'),]
  chem_tf <- chem_data[which(chem_data$Type=='TF'),]
  chem_tp <- chem_data[which(chem_data$Type=='TP'),]
  #merge on site, date, compound
  chem_merge <- merge(chem_sf,chem_tf,by=c("Compound","Site","Date"))
  chem_merge <- merge(chem_merge, rainwater, by="Date")
  #1 plot by rainfall (metolachlor and tebuconazole)
  plot(chem_merge$amount_inches, chem_merge$Conc.x, main=paste(chem,'stemflow'))
  #1 plot by date (all together, with 3 different symbols/colors for site and filled/not filled for sf v tf)
  plot(chem_merge$Date, chem_merge$Conc.y, title=paste(chem,'throughfall'))
  plot(chem_merge$Date, chem_merge$Conc.x, main='stemflow')
  #1 stacked barplots by month for all sites for tifton ponds
  
  #conc.x is sf; conc.y is tf
  #we want to test if stemflow is greater than throughfall (x > y)
  x_nas <- sum(is.na(chem_merge$Conc.x))
  y_nas <- sum(is.na(chem_merge$Conc.y))
  x_n <- length(is.na(chem_merge$Conc.x))
  y_n <- length(is.na(chem_merge$Conc.y))
  x_dets <- x_n - x_nas
  y_dets <- y_n - y_nas
  if(x_dets>0){
    if(x_nas==0){
      x_min <- min(chem_merge$Conc.x,na.rm=TRUE)
    }else{
      x_min=0
    }
    x_max <- max(chem_merge$Conc.x,na.rm=TRUE)
  }else{
    x_min <- 0
    x_max <- 0
  }
  if(y_dets>0){
    if(y_nas==0){
      y_min <- min(chem_merge$Conc.y,na.rm=TRUE)
    }else{
      y_min=0
    }
    y_max <- max(chem_merge$Conc.y,na.rm=TRUE)
  }else{
    y_min <- 0
    y_max <- 0
  }
  ### statistical analysis where all 3 locations are grouped
  #colnames(chem_merge)
  print(paste(chem,'stemflow detection frequency = ',x_dets,'/',x_n))
  print(paste(chem,'stemflow concentration range =(',x_min,',',x_max,')'))
  print(paste(chem,'throughfall detection frequency = ',y_dets,'/',y_n))
  print(paste(chem,'throughfall concentration range =(',y_min,',',y_max,')'))
  
  #change NAs to zeroes for conparisons and the Wilcox test
  chem_merge$Conc.x[is.na(chem_merge$Conc.x)] <- 0
  chem_merge$Conc.y[is.na(chem_merge$Conc.y)] <- 0
  n_comparisons <- length(chem_merge$Conc.x)
  n_tf_greater <- sum(chem_merge$Conc.y > chem_merge$Conc.x)
  n_ties <- sum(chem_merge$Conc.y == chem_merge$Conc.x)
  n_sf_greater <- sum(chem_merge$Conc.x > chem_merge$Conc.y)
  comparisons <- paste('(throughfall_greater, ties, stemflow_greater) = ('
              ,n_tf_greater,',',n_ties,',',n_sf_greater,')')
  print(comparisons)
  
  #do the test for all sites
  if((x_dets+y_dets)>4){
    print(knitr::kable(chem_merge))
    chem_wilcox <- wilcox.test(chem_merge$Conc.x, chem_merge$Conc.y, 
                               alternative = 'greater', paired = TRUE, exact = FALSE)
    max_conc <- max(x_max,y_max)
    max_conc_vector <- seq(0,max_conc,20)
    print(chem_wilcox)
    plot(chem_merge$Conc.x, chem_merge$Conc.y,xlim=c(0,max_conc),ylim=c(0,max_conc),
         xlab='Stemflow Concentration',ylab='Throughfall Concentration',
         main=paste(chem,'(WRS p-value = ',round(chem_wilcox$p.value,4),')'),
         sub = comparisons)
    abline(0,1,col='red')
  }else{
    print(paste(chem,'does not have enough detected values for wilcox test'))
  }
  ### statistical analysis for each location, not grouped
  #do the test for each sites
}
```


  
global analyses
====================================
Two values with high concentrations are not shown on the all tifton data figure.

```{r eval=TRUE, echo=TRUE}
print('=======================================================')
print('global')
tifton_sf <- tifton[which(tifton$Type=='SF'),]
tifton_tf <- tifton[which(tifton$Type=='TF'),]
#merge on site, date, compound
tifton_merge <- merge(tifton_sf,tifton_tf,by=c("Compound","Site","Date"))
#change NAs to zeroes for comparisons and the Wilcox test
tifton_merge$Conc.x[is.na(tifton_merge$Conc.x)] <- 0
tifton_merge$Conc.y[is.na(tifton_merge$Conc.y)] <- 0
n_tf_greater <- sum(tifton_merge$Conc.y > tifton_merge$Conc.x)
n_ties <- sum(tifton_merge$Conc.y == tifton_merge$Conc.x)
n_sf_greater <- sum(tifton_merge$Conc.x > tifton_merge$Conc.y)
comparisons <- paste('(throughfall_greater, ties, stemflow_greater) = ('
            ,n_tf_greater,',',n_ties,',',n_sf_greater,')')
print(comparisons)

tifton_wilcox <- wilcox.test(tifton_merge$Conc.y, tifton_merge$Conc.x, 
                               alternative = 'two.sided', paired = TRUE, exact = FALSE)
x_max <- max(tifton_merge$Conc.x)
y_max <- max(tifton_merge$Conc.y)
max_conc <- max(x_max,y_max)
max_conc_vector <- seq(0,max_conc,20)
print(tifton_wilcox)
plot(tifton_merge$Conc.x, tifton_merge$Conc.y, 
    xlim=c(0,4),ylim=c(0,4),
    xlab='Stemflow Concentration',ylab='Throughfall Concentration',
    main=paste('All Tifton Data (WRS p-value = ',round(tifton_wilcox$p.value,4),')'),
    sub = comparisons)
abline(0,1,col='red')
```

date analyses
====================================
```{r eval=TRUE, echo=TRUE}
#break down by day
#rainwater
View(rainwater)
#tifton[which(tifton$Compound==chem),]
tifton_sf <- tifton[which(tifton$Type=='SF'),]
dim(tifton_sf)
tifton_tf <- tifton[which(tifton$Type=='TF'),]
dim(tifton_tf)
tifton_tp <- tifton[which(tifton$Type=='TP'),]

#merge on site, date, compound
tifton_merge <- merge(tifton_sf,tifton_tf,by=c("Compound","Site","Date"))
dim(tifton_merge)
tifton_merge <- merge(tifton_merge, rainwater, by="Date")
dim(tifton_merge)
unique_dates <- unique(tifton_merge$Date)

for(date in unique_dates){
  tifton_merge_date <- tifton_merge[which(tifton$Date==date),]
  #change NAs to zeroes for comparisons and the Wilcox test
  tifton_merge_date$Conc.x[is.na(tifton_merge_date$Conc.x)] <- 0
  tifton_merge_date$Conc.y[is.na(tifton_merge_date$Conc.y)] <- 0
  print(date)
  #print(tifton_merge_date)
  n_tf_greater <- sum(tifton_merge_date$Conc.y > tifton_merge_date$Conc.x)
  n_ties <- sum(tifton_merge_date$Conc.y == tifton_merge_date$Conc.x)
  n_sf_greater <- sum(tifton_merge_date$Conc.x > tifton_merge_date$Conc.y)
  comparisons <- paste(date,'(throughfall_greater, ties, stemflow_greater) = ('
              ,n_tf_greater,',',n_ties,',',n_sf_greater,')')
  print(comparisons)
  
}

  
```


regressing versus rainfall amounts
====================================
We want to look at how the amount of rainfall influences the concentrations in stemflow and throughfall, in
```{r eval=TRUE, echo=TRUE}
print('=======================================================')
print('global')


```


permutation stat testing
====================================

For many of the chemicals, some/most observations have identical/not detected values, hence the warning message thrown by R: unique ranks cannot be computed for all observations in the Wilcox test. Since there are ties, this prevents the computation of an exact p-value. With high numbers of tied data, Wilcoxon tests are not trustworthy. Moreover, in most non-parametric tests we assume that the sampled populations are symmetric and have the same dispersion or shape, which is not verifiable here.

A permutation test may be more appropriate in this case, for example permTS (perm), pperm (exactRankTests), or the coin package. Although these approaches do not take advantage of the paired design.

```{r eval=TRUE, echo=TRUE}

```

results
====================================
Mixed bag at the moment. Overall inference on paired data shows that throughfall concentrations exceed stemflow concentration 170 times versus 92 of the opposite exceedance. This corresponds to a p-value of 0.0019 with the Wilcoxon Rank Sum test. 

There will be complaints from reviewers about the unequal volumes of samples for the two media, we need to be clear that the collection time is the same for these paired comparisons.

I guess we can also break out how often the exceedances are a detect versus a non-detect as opposed to a detect versus detect comparison. Can we assume that the actual detection limit is lower for the higher volume stemflow samples? (meaning that the volumes may bias the analyses towards stemflow exceedances and not throughfall).

Stemflow concentrations probably plagued by dilution versus first flush issues. We may want to think about how to use sample volume or observed rainfall totals for these dates to look at this and separate small from large rain events.

Having detection limits will help us characterize the magnitude of the difference between stemflow and throughfall (for example, ratio calculations that are not possible by treating non-detects as zeroes).